---
title: "Pt1"
author: "Emanuele Iaccarino"
date: "2025-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(expm)  # for matrix power

# transition matrix
P <- matrix(c(
  0.135, 0.794, 0.071,
  0.461, 0.470, 0.069,
  0.485, 0.420, 0.095
), nrow = 3, byrow = TRUE)

# Compute the invariant distribution π
# Solve π such that π %*% P = π and sum(π) = 1
A <- t(P) - diag(3)
A <- rbind(A, c(1, 1, 1))
b <- c(0, 0, 0, 1)
pi <- solve(t(A) %*% A) %*% t(A) %*% b  # least squares solution

# Compute P^4
P4 <- P %^% 4

# Extract P^4[3,1] (from state 3 to state 1 in 4 steps)
prob_3_to_1_in_4_steps <- P4[3, 1]

cat("Invariant distribution π:\n")
print(pi)
cat("\nP^4:\n")
print(P4)
cat("\nP^4[3,1] (probability from state 3 to 1 in 4 steps):\n")
print(prob_3_to_1_in_4_steps)

```

Invariant distribution π confirms the chain is ergodic and has a unique stationary distribution.

Each row still sums to ~1, confirming it's a valid transition matrix even after multiple steps.

The probability that the Markov chain is in state 1 at time t+4 given that it is in state 3 at time t is:

0.3442953


```{r}
set.seed(22)
n_steps <- 10000

states <- numeric(n_steps)
states[1] <- 1  # Start at state 1

# Simulate the chain
for (t in 2:n_steps) {
  current_state <- states[t - 1]
  probs <- P[current_state, ]
  states[t] <- sample(1:3, size = 1, prob = probs)
}

# Approximate stationary distribution
estimated_pi <- prop.table(table(factor(states, levels = 1:3)))

cat("Estimated stationary distribution after 10,000 steps:\n")
print(estimated_pi)

```
The empirical stationary distribution after 10,000 steps is very close to the theoretical invariant distribution:

This confirms that:

- The Markov chain converges to its stationary distribution.

- The approximation is solid even starting from a non-stationary initial state (state 1), thanks to the long horizon (10,000 steps).

```{r}
# Approximate expected value under stationary distribution
expected_value_estimate <- mean(states)
cat("Estimated expected value under stationary distribution:\n")
print(expected_value_estimate)
```
```{r}
# Compute invariant distribution
A <- t(P) - diag(3)
A <- rbind(A, c(1, 1, 1))
b <- c(0, 0, 0, 1)
pi <- solve(t(A) %*% A) %*% t(A) %*% b

least_state <- which.min(pi) # Least probable state
least_prob <- min(pi) # its probability

# Joint probability P(X_t = 1 and X_{t+1} = 1)
# joint probability of observing state 1 at two consecutive time steps
joint_prob <- pi[1] * P[1, 1]

cat("Least probable state:\n")
print(least_state)
cat("Probability of least probable state:\n")
print(least_prob)
cat("P(X_t = 1 and X_{t+1} = 1):\n")
print(joint_prob)

```

