---
title: "Pt2"
author: "Emanuele Iaccarino"
date: "2025-06-03"
output: html_document
---

```{r}
# Clean current environment and load workspace
rm(list = ls())
load("loyn_data_analysis.RData")

# Summarize JAGS output
summary_stats <- loyn_mod1$BUGSoutput$summary # matrix containing posterior summaries for all parameters

# Extract effective sample sizes
ess <- summary_stats[, "n.eff"] # for each parameter

# Identify parameter with the lowest ESS
lowest_ess_param <- names(which.min(ess)) # thhe function return index so we retrieve name
lowest_ess_value <- min(ess)

cat("Parameter with the lowest effective sample size:\n")
print(lowest_ess_param)
cat("Effective sample size:\n")
print(lowest_ess_value)
```
There was no need actually cause the parameters and effective sample sizes were visible in the next question :(, anyway always worth to check and give robustness to our answer
```{r}
# Extract MCMC samples from the JAGS model
samples_array <- loyn_mod1$BUGSoutput$sims.array

# Check dimension: iterations × chains × parameters
# dim(samples_array)

# Extract samples for beta[3] from chain 1
beta3_chain1 <- samples_array[, 1, "beta[3]"]

# Compute MCSE as sd / sqrt(n)
n <- length(beta3_chain1)
mcse_beta3 <- sd(beta3_chain1) / sqrt(n)

# Print result
cat("Monte Carlo standard error for beta[3] (chain 1 only):\n")
print(mcse_beta3)
```
The Monte Carlo standard error measures the uncertainty in the estimate of the posterior mean due to finite sampling
This is the standard error of the estimate of the posterior mean of beta[3] using only the 3,000 draws from chain 1.
A small MCSE implies that the posterior mean is estimated with high precision from that chain.

To determine how many predictors have a significant effect on the outcome variable (ABUND), we check the 95% credible intervals (CI) for each beta[j]. If the 95% CI does not include 0, then the effect is considered statistically significant in the Bayesian sense.

Only 1 predictor (AREA, beta[2]) has a 95% credible interval that does not contain 0 so it is the only significant predictor.



The 95% credible interval (ET = equal-tail interval) for beta[2] (AREA) is:
CI 95% =[0.222,0.551]

This interval:

- Does not include 0 

- Both endpoints are positive 


Forest plot shows the posterior means and 95% credible intervals for all parameters

It allows to inspect which predictors are statistically significant:

- If the credible interval does not include 0, the predictor is significant.

- If the interval crosses 0, it is not significant.

```{r}
mcmc_chains <- as.mcmc(loyn_mod1)  
combined_matrix <- as.matrix(mcmc_chains)

# Extract only the columns corresponding to beta[1] through beta[7]
beta_columns <- grep("^beta\\[", colnames(combined_matrix), value = TRUE)
beta_samples <- combined_matrix[ , beta_columns ]

# First, compute point estimates (posterior means) for each beta:
beta_means <- apply(beta_samples, 2, mean)

# Create the plot and save to PDF:
pdf("plot.pdf", width = 6, height = 4)
par(mar = c(5, 6, 4, 2))  # margin adjustments: (bottom, left, top, right)

# Plot an empty canvas:
plot(
  x = beta_means,
  y = seq_along(beta_columns),
  xlim = range(c(ci_lower, ci_upper)),
  ylim = c(0.5, length(beta_columns) + 0.5),
  yaxt = "n",
  xlab = "Posterior Mean and 95% Credible Interval",
  ylab = "",
  main = "95% Credible Intervals for Beta Parameters",
  pch = 16
)

# Add custom y-axis labels:
axis(
  side = 2,
  at = seq_along(beta_columns),
  labels = beta_columns,
  las = 1
)

# Add horizontal error bars for each beta:
for (i in seq_along(beta_columns)) {
  seg_x0 <- ci_lower[i]
  seg_x1 <- ci_upper[i]
  seg_y  <- i
  segments(seg_x0, seg_y, seg_x1, seg_y, lwd = 2)
}

# Add a vertical line at zero:
abline(v = 0, lty = 2, col = "darkgray")

dev.off()

cat("Forest plot of 95% credible intervals saved as 'plot.pdf'.\n")
```
