---
title: "SDS2"
author: "Emanuele Iaccarino"
date: "2024-10-10"
output: html_document
---
# Pseudo Random Number Generators (PRNGs)

A pseudo-random number generator (PRNG) is a deterministic algorithm that generates numbers that appear to be random but are actually produced by a fixed sequence of operations. PRNGs are not truly random because they rely on an initial seed value, and the entire sequence of "random" numbers is determined by this seed.

The goal of a PRNG is to generate a sequence:
𝑢0,𝑢1=𝑓(𝑢0),𝑢2=𝑓(𝑓(𝑢0)),…,𝑢𝑛=𝑓(𝑢𝑛−1)
that statistically behaves like independent, identically distributed (iid) random numbers from the uniform distribution on (0,1).

Paradox: PRNGs are deterministic, meaning the same seed will always produce the same sequence. However, the sequence should resemble a random sequence as closely as possible, despite being generated by a fixed rule.

### LCG Function
```{r}
source("Func10.10.R")
'''
# Linear Congruential Generator with the ANSI C default (Saucier, 2000)
lcg <- function(n = 100, M = 2^32, a = 12345, b = 1103515245, u0 = as.numeric(Sys.time()) * 100) {
  out <- rep(NA, n)  # Allocate output
  # Main loop
  for (i in 1:n) {
    u0 <- (a + b * u0) %% M
    out[i] <- u0 / (M + 1)
  }
  return(out)
}
'''
```
# Basic Testing
```{r}
# Basic Testing


lcg # give back the body of the function


lcg(42) # cool, numbers between 0 and 1


# Do they look like IID from a Unif(0,1)?
```
##### 1 Collectively from a Unif(0,1)?
```{r}

?hist

# First try
mytry <- lcg(42) # n = 50
hist(mytry, prob = T)
curve( dunif(x, 0, 1), add = T, lwd = 4, col = "blue")


# Pump-up the number of samples
mytry2 <- lcg(42, n = 5000 ) # n = 50000
hist(mytry2, prob = T)
curve( dunif(x, 0, 1), add = T, lwd = 4, col = "blue")


# Change the default and see what happens
mytry3 <- lcg(42, 
                M = 2^10,
                n = 5000 ) # n = 50000
hist(mytry3, prob = T)
curve( dunif(x, 0, 1), add = T, lwd = 4, col = "blue")
```

##### 2 Independent? | Qualitative Analysis
```{r}
# 2 # Independent? | Qualitative Analysis


# Traceplot!
plot(mytry2, pch = ".")
plot(mytry2[1:50], type = "l", col = rgb(1,0,0,.3), lwd = 4)
# Traceplot will be everywhere in your future MCMC life


# Lag-1 scatterplot
n <- length(mytry2)
plot(mytry2[1:(n-1)], mytry2[2:n], pch = ".")


# Lag-2 scatterplot in the notes


# Lag-1 scatterplot
n <- length(mytry3)
plot(mytry3[1:(n-1)], mytry3[2:n], pch = ".") # BAD!


# Industrial standard -----------------------------------------------------

?set.seed

```

### Quantile Transform
```{r}
# Exponential distro example | we do have a generator available in R
?rexp


# Quantile function for the exponential model

?myqexp
par(mfrow = c(2,1))
curve(dexp(x, rate = .5), 0, 5, lwd = 5, col = "green",
      ylab = "Density")
curve(myqexp(x, alpha = .5), 0, 1, 
      lwd = 5, col = "blue",
      ylab = "Quantile")
par(mfrow = c(1,1))


# Apply the quantile function to the Unif(0,1) PRNG


y <- myqexp(mytry2, alpha = .5)


# Check if they're collectively behaving like a sample (IID)
# from Exp(0.5)


hist(y, prob = T, breaks = 150) # Better pump-up the resolution of your histo
curve(dexp(x, rate = .5), add = T, col = "green", lwd = 4,
      n = 1000)

```

# Empirical example
The Linear Congruential Generator is a specific type of PRNG. It generates a sequence of integers using the formula:
𝑥𝑛+1=(𝑎+𝑏⋅𝑥𝑛) mod𝑀
The constants 𝑎and 𝑏are parameters that control the behavior of the generator, and 𝑀 is the modulus. For properly chosen parameters, the sequence has a period of 𝑀 before repeating.

```{r}
# We can also demonstrate how "random" the LCG is by plotting a sample generation in 3D.
#install.packages("plot3D","animation")
library(plot3D)
library(animation)

n_seq <- c(3, 10, 100, 500, 1000, 2000, 5000, 10000, 20000)

saveGIF({
  for (i in 1:length(n_seq)) {
    x <- lcg(n_seq[i])
    y <- lcg(n_seq[i])
    z <- lcg(n_seq[i])
    scatter3D(x, y, z, colvar = NULL, pch = 20, cex = .5, theta = 20, main = paste("n =", n_seq[i]))
  }
}, movie.name = "lcg.gif")

```
### Poor parameters choice, Empirical result
```{r}
# What happens with poor choices of parameters?
saveGIF({
  for (i in 1:length(n_seq)) {
    x <- lcg(n_seq[i], a = 1, b = 1229, M = 2^11)
    y <- lcg(n_seq[i], a = 1, b = 1229, M = 2^11)
    z <- lcg(n_seq[i], a = 1, b = 1229, M = 2^11)
    scatter3D(x, y, z, colvar = NULL, pch = 20, cex = .5, theta = 20, main = paste("n =", n_seq[i]))
  }
}, movie.name = "lcg_poor.gif")

```
The LCG produces a sequence that is not truly random.
The sequence may repeat or show patterns due to a small modulus 𝑀 (the sequence repeat itself after 𝑀 steps ) and inadequate values for 𝑎 and 𝑏 (points will not be uniformly scattered but they will show some pattern or cluster).

# Generating 1D Random variable

### Continuous Random Variable Simulation
```{r}
# Continuous random variable
fx <- function(x) {
  3 * (x^2) * ((x >= 0) * (x <= 1))
}

Fx <- function(x) {
  (x^3) * ((x >= 0) * (x < 1)) + (x >= 1)
}

Qx <- function(p) {
  (p^(1/3)) * (p >= 0) * (p <= 1)
}

# Simulation
set.seed(1234)
M <- 1000 # Simulation size
u <- runif(M)
x <- Qx(u)

# Histogram / PDF
hist(x, prob = TRUE, col = "orchid", border = "white", ylim = c(0, 3), main = "", xlab = "")
rug(x)
curve(fx(x), lwd = 8, col = rgb(0.45, 0.16, 0.69, 0.5), add = TRUE)
legend("topleft", "True PDF", col = rgb(0.45, 0.16, 0.69, 0.5), lwd = 8, bty = "n", cex = 0.7)

# ECDF / CDF
plot(ecdf(x), main = "", xlab = "")
rug(x)
curve(Fx(x), lwd = 8, col = rgb(0.45, 0.16, 0.69, 0.3), add = TRUE)
legend(0, 1, c("Empirical CDF", "True CDF"), col = c("black", rgb(0.45, 0.16, 0.69, 0.5)), lwd = c(1, 8), bty = "n", cex = 0.7)

```

### Discrete Random Variable Simulation
```{r}
# Discrete random variable

# Simulation setup
M <- 1000 # Simulation size
p <- 0.4  # Pr(success) = Pr(X = 1)

# 1. Bernoulli(p) = Binom(p, 1)
x1 <- rbinom(M, size = 1, prob = p)
t1 <- table(x1) / M; t1

# 2. Quantile transform
u <- runif(M)
x2 <- (u > 1 - p) * (u < 1)
t2 <- table(x2) / M; t2

# 3. Generic sampling of a discrete random variable
x3 <- sample(c("Head", "Tail"), size = M, prob = c(1 - p, p), replace = TRUE)
t3 <- table(x3) / M; t3

# Plot
tab <- rbind(t1, t2, t3)
dat <- barplot(tab, ylim = c(0, 1), beside = TRUE, border = "white",
               col = viridis::viridis(3, alpha = 0.5), legend.text = c("X ~ rbinom(p, 1)", "Quantile-transform", "sample()"),
               args.legend = list(border = "white", cex = 0.8, bty = "n"))
text(x = dat, y = c(tab), labels = c(tab), pos = 3, cex = 0.8)

```
