---
title: "Stats"
author: "Emanuele Iaccarino"
date: "2025-01-29"
output: html_document
---

# 20 Dec 2016

```{r}
# to calculate confidence interval
alpha = 0.1
z_score <- qnorm(1 - alpha/2)
```

##### A

```{r}
# Data
xx <- c(1.1, 1.7, 2.0, 9.3, 1.4, 1.2)
# MLE functions
a.hat <- function(x) length(x)/sum(log(x))
se.hat <- function(x) abs(length(x)/sum(log(x)))/sqrt(length(x))
# Evaluate the CI
cat('--- MLE -------------\n')
round(a.hat(xx), 3)
cat('--- SE --------------\n')
round(se.hat(xx), 3)
cat('--- CI(0.99) --------\n')
round(
c(lower = a.hat(xx) - qnorm(1-alpha/2)*se.hat(xx),
upper = a.hat(xx) + qnorm(1-alpha/2)*se.hat(xx)), 3)
```

##### C

```{r}
mle <- a.hat(xx)
# CDF function
cdf = function(x, alpha) (1 - x^(-alpha))*(x >= 1)
# Plot: very few datapoints, endorsing every parametric model is a risky biz!
curve( cdf(x, mle), from = -2, to = 12, lwd = 4, col = "orchid", ylim = c(0,1),
xlab = expression(alpha), ylab = "CDF function")
plot(ecdf(xx), cex = .7, add = T)
grid()
```
```{r}
# Sampling
rquest <- function(n, alpha){
u <- runif(n)
out <- ((u)^(1/alpha))^(-1)
return(out)
}
```

```{r}
# Parametric Boostrap
n <- length(xx) # Sample size
B <- 1000 # Bootstrap size
a.boot <- rep(NA, B) # Init
set.seed(123)
for(b in 1:B){
x.boot <- rquest(n, a.hat(xx))
a.boot[b] <- a.hat(x.boot)
}
# Bias via Parametric Bootstrap
cat('--- PBoot / Bias ----\n')
round(mean(a.boot) - a.hat(xx), 3)
# SE via Parametric Bootstrap
cat('--- PBoot / SE ------\n')
round(sd(a.boot), 3)
# Parametric bootstrap CI
cat('--- PBoot CI(0.90) --\n')
round(
c(lower = a.hat(xx) - qnorm(1-0.01/2)*sd(a.boot),
upper = a.hat(xx) + qnorm(1-0.01/2)*sd(a.boot)), 3)
```
# 2020 February 13 Exam-03 - Part I (Probability)

##### A
Theory Question:
Independence between events: Briefly explain the different types of independence between two or more events and their relations.

```{r}
# Pairwise indipendence (between 2 events)
Two events A and B are independent if the Prob of their intersection is equal to the product of the prob of each event. 
# P(A intersect B) = P(A) * P(B)
An example is to toss a fair coin, where the prob of tossing a Head on first event is equal to the one of tossing Head also on the second attempt, so the prob of tossing Head twice is just their product
# P(A intersect B) = 1/2 * 1/2 = 1/4 and it match

#  Mutual (Joint) Independence
We consider more than 2 events in this case.
Back to the Coin problem, if we want to obtain head a third time we can follow a similar approac 
# P(A intersect B intersect C) = (1/2)^3
Mutual independence implies pairwise independence, but pairwise independence does not necessarily imply mutual independence.

# Conditional Independence
so they are independent only given a third event C

# P(A interesect B∣C)=P(A∣C)P(B∣C)
his means that if we know C has occurred, then the probability of (A intersect B) is just the product of their individual probabilities given C
```

You enter a League of Legends tournament where your probability of winning a game is 0.3 against half the players (call them Type 1), 0.4 against a quarter of the players (call them Type 2) and 0.5 against the remaining quarter of the players (call them Type 3).
You play a game against a randomly chosen opponent.

1. What is the probability of winning the game?
```{r}
p1 = 1/2 # P (T_1)
p2 = 1/4 # P (T_2)
p3 = 1/4 # P (T_3)

P_W_given_T1 = 0.3 # P (W|T_1)
P_W_given_T2 = 0.4 # P (W|T_2)
P_W_given_T3 = 0.5 # P (W|T_3)

# Using Law of total probabilities
PW = P_W_given_T1 * p1 + P_W_given_T2 * p2 + P_W_given_T3 * p3
PW
```
Suppose you win. What is the probability that you had an opponent of Type 1?
```{r}
# P(T_1|W) using Bayes Theorem
PT1_W = P_W_given_T1 * p1 / PW
PT1_W
```
##### B

```{r}
Fx = function(x) 3*(4/3 - 2/x)*(x >= 3/2)*(x < 2) + (x >= 2)
curve(Fx(x), from =1, to = 2.5, n = 300)
```

##### C
Theory Question:
Mean vs Median: contrast these two distribution summaries in terms of their main properties.

```{r}
The mean is the sum of all the observation divided by the total number of observation.
It consider all the data points so it is affected by outliers.

The median is the middle point of the ordered dataset, if the number of data points is odd, the median is the middle value, instead if the number is even, the median is the average between the two middle values.
It is more robust for outliers, so we prefer median over mean in skewed distribution, instead for symmetric distribution we prefer the mean since computationally is less expensive.
```

```{r}
ell <- 1  # Stick length
n <- 10^6  # Number of simulations

# Generate X1 (first break point)
X1 <- runif(n, min = 0, max = ell)

# Generate X2 (second break point, inside X1)
X2 <- runif(n, min = 0, max = X1)

# Compute expected value from simulation
E_X2_simulated <- mean(X2)

E_X1 = ell / 2 # we break at a random point, so mean is the avg

# Compute theoretical expectation
E_X2_theoretical <- E_X1 * 1/2 # we divide once again by half 

round(E_X2_simulated, 3)
round(E_X2_theoretical, 3)

```


# 2020 February 13 Exam-03 - Part II (Statistics)

##### A
Theory Question:
Point Estimation: describe a general decision theoretical framework to evaluate/compare point estimators. Then provide a specific example.

```{r}
1. We define an estimator:
  is a function of observed data that provides a single best guess for an    unknown population parameter
#Ex:
To estimate the population mean mu, our estimator will be the sample mean mu_hat so we do the mean of the sum of all the observed data points
2. Loss function:
  cost of an estimation error
# Ex:
We can use squared error loss to penalize more larger errors
Or we can use absoluted error loss to deal with outliers
3. Risk function:
  Expected loss under the true population parameter
# Ex:
For squared error loss:
# R(θ_hat,θ) = E[( θ_hat − θ)^2 ] 
so the mean of the difference between population and sample parameter (squared in this case)
# = Var(θ_hat ) + (Bias (θ_hat) )^2
from the bias-variance decomposition

4. Comparing Estimators
We use Mean Squared Error (MSE)
# MSE( θ_hat )=E[( θ_hat −θ)^2 ]
# = Var( θ_hat )+Bias( θ_hat )^2
A better estimator has a lower MSE.

```

```{r}
func0 = function(mu, n) mu^2 * (1/(n+1)^2) + n/(n+1)^2
funcXn = function(mu, n) 1/n
n = 100

#n = 10 works with any value of n
mu_values = seq(-2, 2, length.out = n)

y_values0 = func0(mu_values, n)
plot(y_values0)
abline(h = funcXn(mu_values, n))
# neither dominates the other
```

##### B

The dataset kill_bill.RData contains the times (in minutes from the beginning) at which a swear-word/blasphemy is said during the first hour of Quentin Tarantino’s Kill Bill saga.
For the sake of this exercise, assume they are independent realizations of a Gamma random variable X with rate parameter β and shape parameter α both unknown.
with E(X) = α/β and Var(X) = α/β^2
```{r}
kill_bill <- read.csv2("C:/Users/emanu/Downloads/kill_bill.csv", sep="")
kill_bill$time_in_minutes <- as.numeric(kill_bill$time_in_minutes) # idk why was given in character type
View(kill_bill)
```
```{r}
X <- kill_bill$time_in_minutes

# Compute sample statistics
X_bar <- mean(X)      # Sample mean
X_var <- var(X)       # Sample variance

# Compute estimates
alpha_hat <- round((X_bar^2 / X_var),3) # shape
beta_hat <- round((X_bar / X_var),3) # rate
# even easier, no need to create functions
```

```{r}
alpha_hat
beta_hat
```

Is it a good fit? Detail your answer using suitable plots and stats.
```{r}
hist(kill_bill$time_in_minutes, prob = T) # actual data
curve( dgamma(x, shape = ab_mom[1], rate = ab_mom[2]), add = T) #gamma fit
```
```{r}
#PDF
hist(kill_bill$time_in_minutes, prob = T) # actual data
curve( dgamma(x, shape = alpha_hat, rate = beta_hat), add = T, col = "red") #gamma fit
# T allows to compare on same graph actual data and gamma fit
```
```{r}
# CDF
curve( pgamma(x, shape = alpha_hat, rate = beta_hat), from = min(kill_bill), to = max(kill_bill), col = "red")

plot(ecdf(kill_bill$time_in_minutes), add = T)
```

```{r}
# Q-Q plot
n <- length(kill_bill$time_in_minutes)

qqplot(
  qgamma(ppoints(n), shape = alpha_hat, rate = beta_hat),  # Theoretical quantiles
  kill_bill$time_in_minutes,  # Observed data
)
qqline(
  kill_bill$time_in_minutes,
  distribution = function(p) qgamma(p, shape = alpha_hat, rate = beta_hat),
  col = "red"
)

```
Comment: Since I didn't had access to the original data of the Exam I asked ChatGPT to create my same, he uses a poisson distribution with a rate of 1.5 swear word per minute in a 60 minutes time frame. Nonetheless we managed somehow to approximate the distribution in a valuable way, results are encouraging

##### C

use the parametric bootstrap to provide a 95% confidence interval for the population mean.
```{r}
B = 1000
mu_boot = rep(NA, B) # Initialize
set.seed(1234)
for (b in 1:B){ # for each Bootstrap b
  x_boot = rgamma(n, shape = alpha_hat, rate = beta_hat) # Generate bootstrap sample
  mu_boot[b] = mean(x_boot) # Save mean under index b
}
# mu_boot shows mean for each bootstrap sample
# Confidence intervals
alpha <- 0.05 
zz <- -qnorm(alpha/2) # zscore value for alpha parameter (95% of normal distribution)
# also qnorm(1-(alpha/2))
# Now we calculate critical value 
# CI (confidence interval) = [mu-z(alpha/2)*SE,mu+z(alpha/2)*SE]
# mu is sample mean, z(alpha/2) is zz previously calculated, SE is standard error, that we estimate as the standard deviation of all the Bootstap samples
low = round(mean(kill_bill$time_in_minutes) - zz*sd(mu_boot),3) # Lower Bound
up = round(mean(kill_bill$time_in_minutes) + zz*sd(mu_boot),3) # Upper Bound

low
up
```

Double check the previous result by building a pivotal, nonparametric bootstrap based, confidence interval at the same level for the population mean. Compare the two results.

Theory:
parametric bootstrap assumes a Gamma distribution for resampling, 
nonparametric bootstrap simply resamples directly from the observed data without assuming an underlying distribution

pivotal method constructs the confidence interval using the empirical quantiles of the bootstrap means, adjusting the sample mean accordingly
```{r}
mu_nboot = rep(NA, B) # Initialize a new one
for (b in 1:B) {
  x_nboot = sample(kill_bill$time_in_minutes, replace = TRUE)  # Resample from observed data
  mu_nboot[b] = mean(x_nboot)  # Compute mean of bootstrap sample
}
# Get bootstrap quantiles
nlow = round( quantile (mu_nboot, (alpha/2)),3) # Lower Bound
nup = round( quantile (mu_nboot, (1 - alpha/2)),3) # Upper Bound

nlow
nup
```
Comment:
Quite similar to each other, since non parametric Bootstrap is data driven it doesn't assume any specific distribution so it's more robust when data distribution is unknown, Parametric Bootstrap ofc is more efficient if the Gamma distribution assumption is correct


