---
title: "Exams"
author: "Emanuele Iaccarino"
date: "2025-01-30"
output: html_document
---

# Feb 2017

#####A

```{r}
# Define probabilities for each girl
P_A <- 2/9
P_B <- 4/9
P_C <- 2/9
P_D <- 1/9

# Define conditional probabilities for Tullia's approval
P_T_given_A <- 0.3
P_T_given_B <- 0.0
P_T_given_C <- 0.3
P_T_given_D <- 0.6

# Compute P(T) using the Law of Total Probability
P_T <- (P_T_given_A * P_A) + (P_T_given_B * P_B) + (P_T_given_C * P_C) + (P_T_given_D * P_D)

# Compute P(A | T) using Bayes' Theorem
P_A_given_T <- (P_T_given_A * P_A) / P_T

P_T
P_A_given_T
```
##### C


```{r}
# Data
mu_A <- 85
sigma_A <- 10
mu_B <- 90
sigma_B <- 16
rho <- 0.8 #p coefficient
X_A_given <- 80 #score given
```

If the student’s score on test A is 80, what is the probability that her score on test B will be higher than 90?
```{r}
# Compute conditional mean and variance using cond normal distribution formula
mu_B_given_A <- mu_B + (sigma_B / sigma_A) * rho * (X_A_given - mu_A)
sigma_B_given_A <- sqrt((1 - rho^2) * sigma_B^2)

# Compute probability P(X_B > 90 | X_A = 80)
p_xb_gt_90 <- 1 - pnorm(90, mean = mu_B_given_A, sd = sigma_B_given_A)

mu_B_given_A
sigma_B_given_A
p_xb_gt_90
```
If a student is chosen at random, what is the probability that her score on test A will be higher than his score on test B?

```{r}
# Compute mean and variance of Delta
mu_Delta <- mu_A - mu_B
sigma_Delta <- sqrt(sigma_A^2 + sigma_B^2 - 2 * (sigma_A * sigma_B * rho))

# Compute probability P(X_A > X_B)
p_xa_gt_xb <- 1 - pnorm(0, mean = mu_Delta, sd = sigma_Delta)

mu_Delta
sigma_Delta
p_xa_gt_xb

```
If a student is chosen at random, and her score on test B is 100, what predicted value of her score on test A has the smallest mse and what is the value of this minimum mse?

```{r}
X_B_given <- 100 # observed value

# Compute conditional mean
mu_A_given_B <- mu_A + (sigma_A / sigma_B) * rho * (X_B_given - mu_B)

# Compute minimum MSE (Variance of conditional distribution)
mse_min <- (1 - rho^2) * sigma_A^2

mu_A_given_B
mse_min

```
# Jan 2017

##### A
```{r}
# Data
coins = 3
p1 = 0 # P(H|C_1)
p2 = 3/4 # P(H|C_2)
p3 = 1 # P(H|C_3)
```

Suppose that one coin is selected at random from the box and that, when it is tossed once, a Head is obtained. What is the posterior probability that the i^th coin was selected for i ∈ {1, 2, 3}?
```{r}
p = 1/3
P_H = p1 * p + p2 * p + p3 * p # Law of total probabilities

# Posterior probabilities using Bayes rule
P_C1_given_H = p1 * p / P_H
P_C2_given_H = p2 * p / P_H
P_C3_given_H = p3 * p / P_H
```
If the same coin were tossed again, what would be the probability of obtaining another Head?
```{r}
# P(H_2|H_1) = ?
P_H2_given_H1 =  P_C1_given_H * p1 + P_C2_given_H * p2 + P_C3_given_H * p3
```
3. If a Tail had been obtained on the first toss of the selected coin and the same coin were tossed again,
what would be the probability of obtaining a Head on the second toss?
```{r}
#P(Ci|H_1c) = ?
# p of obtaining Tail on the first throw
P_Hc_given_C1 = 1 - p1
P_Hc_given_C2 = 1 - p2
P_Hc_given_C3 = 1 - p3
P_C = P_Hc_given_C1 * p + P_Hc_given_C2 * p + P_Hc_given_C3 * p

P_C1_given_Hc = P_Hc_given_C1 * p / P_C
P_C2_given_Hc = P_Hc_given_C2 * p / P_C
P_C3_given_Hc = P_Hc_given_C3 * p / P_C

P_H2_given_H1c =  P_C1_given_Hc * p1 + P_C2_given_Hc * p2 + P_C3_given_Hc * p3
```
##### C

```{r}
n_Cs = 0.2
E_Cs = 80
n_Econ = 0.3
E_Econ = 76
n_Sts = 0.5
E_Sts = 84
```

```{r}
E_School = (n_Cs * E_Cs) + (n_Econ * E_Econ) + (n_Sts * E_Sts)
E_School
```

# January 2024

##### A

```{r}
P_H = 3/4
P_Hc = 1/4

P_X_given_H <- 0.08
P_X_given_Hc <- 0.40

# Compute P(X = 0)
P_X0 <- (P_X_given_H * P_H) + (P_X_given_Hc * P_Hc)
P_H_given_X0 <- (P_X_given_H * P_H) / P_X0
P_H_given_X0
```

```{r}
P_X16_given_H <- 0.20
P_X16_given_Hc <- 0.08

# Compute P(X = 16)
P_X16 <- (P_X16_given_H * P_H) + (P_X16_given_Hc * P_Hc)

P_H_given_X16 <- (P_X16_given_H * P_H) / P_X16
P_H_given_X16
```

```{r}
mu_H <- 15
sigma2_H <- 50

mu_Hc <- 8
sigma2_Hc <- 50

# Compute Normal PDFs at X = 0
P_X0_given_H <- dnorm(0, mean = mu_H, sd = sqrt(sigma2_H))
P_X0_given_Hc <- dnorm(0, mean = mu_Hc, sd = sqrt(sigma2_Hc))

P_H_given_X0_norm <- (P_X0_given_H * P_H) / ((P_X0_given_H * P_H) + (P_X0_given_Hc * P_Hc))
P_H_given_X0_norm
# on denominator we use Law of total probabilities to retrieve P(X=0)

```

```{r}
P_X10_15_given_H <- pnorm(15, mean = mu_H, sd = sqrt(sigma2_H)) - pnorm(10, mean = mu_H, sd = sqrt(sigma2_H))
P_X10_15_given_H
# From table is 0.35 so it doesn't match, Normal assumption is not reasonable
```
##### C

```{r}
# Y_1 = X_1 + X_2
# Y_2 = X_1 - X_2
n <- 10000

# Generate random X1 and X2
X1 <- sample(c(-1,1), n, replace = TRUE)
X2 <- sample(c(-1,1), n, replace = TRUE)

# Compute Y1 and Y2
Y1 <- X1 + X2
Y2 <- X1 - X2

# Compute Pearson correlation
correlation <- cor(Y1, Y2)
# Close to 0 since they are uncorrelated

```

