---
title: "Past Exams"
author: "Emanuele Iaccarino"
date: "2025-03-16"
output: html_document
---
# December 2016 v1
Evaluate the ci you’ve just build on the following dataset and briefly comment on its reliability
```{r}
xx <- c(2.2, 2.7, 1.0, 1.4, 1.1)
a_hat = function(x) {length(x)/sum(log(x))}
se_hat = function(x) {(length(x)/sum(log(x))) / sqrt(length(x))}

a_hat(xx)
se_hat(xx)
# lower
alpha = 0.1
z = qnorm(1 - alpha/2)
a_hat(xx) - z * se_hat(xx)
# upper
a_hat(xx) + z * se_hat(xx)
```
Since we have a small number of data points the reliability is low

Plot the likelihood function L(α | x5) highlithing the mle.
```{r}
L <- function(alpha, x){
n <- length(x)
jnk <- prod(1/x) # productoria
out <- (alpha^n)*(jnk)^(alpha+1)
return(out)
}
```

```{r}
alpha_vals <- seq(0.1, 8, length.out = 100)

# Compute likelihood values
L_vals <- sapply(alpha_vals, function(a) L(a, xx))

# Plot
plot(alpha_vals, L_vals)

# Add MLE estimate as a vertical line
abline(v = a_hat(xx), col = "red")
```
```{r}
f = function(x) {1 / (2*(1 - x)^(1/2))}
res = integrate(f, lower = 0, upper = 1/2)
res
```
# Jan 26 2017 Stats

```{r}
# The data
dat <- c(10.2, 10.0, 8.0, 1.8, 6.5, 4.4, 9.3, 7.6, 10.1, 7.8)
# The parameter <tau> as a function of <psi>
tau <- function(psi) (2 - pi/2)*(psi)
# The MLE for <psi>
mle <- function(x) (sum(x^2)/(2*length(x)))
# Evaluate the MLE's & take a look
psi_squared.ML <- mle(dat)
tau.ML <- tau(psi_squared.ML)
c(psi_squared.ML = psi_squared.ML, tau.ML = tau.ML)
```
```{r}
# Density function
dmy <- function(x, psi) (x/psi^2)*exp(-x^2/(2*psi^2))*(x >= 0)
# Plot
psi.seq = c(1, 5, 10)
curve(dmy(x,psi.seq[1]), from = 0, to = 30, n = 300, col="red")
curve(dmy(x,psi.seq[2]),add = TRUE, col = "purple")
curve(dmy(x,psi.seq[3]), add = TRUE, col = "blue")
# The curves reach the peak at its psi value, anyway the greater, more flat is the curvature
```

```{r}
rmy = function(n, psi) { psi * sqrt(- 2*log(1- runif(n)))}
M = 10000
hist(rmy(M,psi.seq[1]), col="pink", prob = T)
curve(dmy(x,psi.seq[1]),add = TRUE, col="red")
hist(rmy(M,psi.seq[2]), col="pink", prob = T)
curve(dmy(x,psi.seq[2]),add = TRUE, col = "purple")
hist(rmy(M,psi.seq[3]), col="pink", prob = T)
curve(dmy(x,psi.seq[3]), add = TRUE, col = "blue")

```
```{r}
psi0 = 5.911721
tau0 <- 15
n <- 100
a <- 0.05
z = qnorm(1 -a/2)
M = 1000

# Define functions
tau <- function(psi) (2 - pi/2) * (psi^2) # defined as population variance
mle <- function(x) sqrt(sum(x^2) / (2 * length(x)))
se.tau <- function(x) (abs(2 * (2 - pi/2) * mle(x))) * (abs(mle(x))) / (2 * sqrt(length(x))) # it uses delta method

# Generate M samples and compute statistics iteratively
psi.mle.vec <- numeric(M)  # Store MLEs
tau.mle.vec <- numeric(M)  # Store tau estimates
tau.se.vec <- numeric(M)   # Store SEs

for (i in 1:M) {
  X <- rmy(n, psi = psi0)  # Generate a single sample
  psi.mle.vec[i] <- mle(X)  # Compute MLE
  tau.mle.vec[i] <- tau(psi.mle.vec[i])  # Compute tau
  tau.se.vec[i] <- se.tau(X)  # Compute standard error
}

# Compute test statistics
W <- abs((tau.mle.vec - tau0) / tau.se.vec)
mean(abs(W) > z)
```
# Feb 2017 Stats

```{r}
xx = c(0.65, 0.48, 0.08, 0.11, 0.80, 0.67, 0.38, 0.63, 0.10, 0.03)
psi = function(x) {- length(x)/sum(log(1-x))}
tau = function(x) {1 / (1+psi(x))}
psi(xx)
tau(xx)
```

```{r}
tau0 = 1
z = qnorm(1-0.05/2)
se.tau = function(x) ( 1/((1 + psi(x))^2) )*(abs(psi(x)))/(sqrt(length(x)))
W = abs((tau(xx) - tau0)/se.tau(xx))
W > z
# We reject the null at \alpha = 0.05
```
Plot the likelihood function L(ψ | x10) highlithing the mle
```{r}
L = function(psi, x){
  n = length(x)
  s = prod(1-x)
  out = psi^n * s^(psi-1)
}
curve(L(x,xx), from=0, to= 4)
segments(psi(xx),0,psi(xx),L(psi(xx),xx)) # MLE
# we start do draw at x = MLE and y = 0 and we draw a straight line since x1 remains the same(MLE) while we extend y until the end of the Likehood function (until it touches)
```

```{r}
# Parametric Bootstrap
# we assumed that data follows a distribution Beta(1, mle)
n = length(xx)
B = 1000
tau.boot <- rep(NA, B) # Init
for(b in 1:B){
x.boot <- rbeta(n, 1, psi(xx)) # bootstrap samples
tau.boot[b] <- 1/(1 + psi(x.boot))
}
bias = mean(tau.boot) - tau(xx)
bias
se = sd(tau.boot)
se
lower = tau(xx) - z *se
lower
upper = tau(xx) + z *se
upper
```
```{r}
# Non parametric Bootstrap
# we do not assume any probabilistic distribution, we rely on empirical resampling from the observed dataset
for (b in 1:B){
  idx <- sample(1:n, replace = T) # resample data with replacement
  x.boot <- xx[idx]
  tau.boot[b] <- 1/(1 + psi(x.boot))
}
bias = mean(tau.boot) - tau(xx)
bias
se = sd(tau.boot)
se
lower = tau(xx) - z *se
lower
upper = tau(xx) + z *se
upper
```

Since both methods produce similar results, we can claim that assuming a beta distribution do not introduce significant bias into our estimates. Since non parametric provided a lower Standard error, the estimates might be slightly more precise.

# Nov 2017 Prob 

```{r}
exact_cov = function(n,k) {- (n/k^2) }
game = function(n,k) {sample(x = 1:k, size = n, replace = T)}
n = 100
print("K=6")
k = 6
M = 10000
games_1 = replicate(M,sum(game(n = 100, k = 6)==1))
games_2 = replicate(M,sum(game(n = 100, k = 6)==2))
s1 = sum(games_1)/M
s2 = sum(games_2)/M
cov(s1,s1)  #aprox
exact_cov(n = 100, k = 6) #est
print("K=12")
k = 12
games = replicate(M,game(n = 100, k = 12))
games_1 = replicate(M,sum(game(n = 100, k = 12)==1))
games_2 = replicate(M,sum(game(n = 100, k = 12)==2))
s1 = sum(games_1)/M
s2 = sum(games_2)/M
cov(s1,s1) #aprox
exact_cov(n = 100, k = 12) #est

```
I surrender

# 20 Dec 2017 Prob

```{r}
# Tried diff value of s, n1, n2
s = 5
n1 =10
n2 = 20
s^2/4 * (1/n1 + 1/n2)
s^2 / (n1+n2)
```

```{r}
x1 = c(0.65, 0.48, 0.08, 0.11, 0.80, 0.67, 0.38, 0.63, 0.10, 0.03)
x2 = c(0.10, 0.86, 0.64, 0.38, 0.57, 0.60, 0.50, 0.08, 0.10, 0.07)

psi = function(x) {-length(x)/sum(log(1-x))}
tau = function(x) {1/(1+(-length(x)/sum(log(1-x))))}
print("x'")
psi(x1)
tau(x1)
print("x''")
psi(x2)
tau(x2)
```
```{r}
se_tau = function(x) {
  psi_ml = -length(x) / sum(log(1 - x))  # MLE for psi
  se_psi_ml = psi_ml / sqrt(length(x))  # Standard error of psi
  se_tau_ml = abs(-1 / (1 + psi_ml)^2) * se_psi_ml  # Delta method
  return(se_tau_ml)
}
alpha1 = 0.1
alpha2 = 0.01
z1 <- qnorm(1 - alpha1/2)
z2 <- qnorm(1 - alpha2/2)
print("x1, z1")
tau(x1) - z1 * se_tau(x1)
tau(x1) + z1 * se_tau(x1)
print("x2, z1")
tau(x2) - z1 * se_tau(x2)
tau(x2) + z1 * se_tau(x2)
print("x1, z2")
tau(x1) - z2 * se_tau(x1)
tau(x1) + z2 * se_tau(x1)
print("x2, z2")
tau(x2) - z2 * se_tau(x2)
tau(x2) + z2 * se_tau(x2)
```
Parametric bootstrap
```{r}
n = length(x1)
B = 1000
tau.boot <- rep(NA, B) # Init
for(b in 1:B){
x.boot <- rbeta(n, 1, psi(x1)) # bootstrap samples
tau.boot[b] <- 1/(1 + psi(x.boot))
}
bias = mean(tau.boot - tau(x1))
print("Boot bias")
bias
se = sd(tau.boot)
print("Boot se")
se
print("Asymptotical se")
se_tau(x1)
print("lower Ci")
lower = tau(x1) - qnorm(1 - alpha2/2)*se
lower
print("upper Ci")
upper = tau(x1) + qnorm(1 - alpha2/2)*se
upper
```
```{r}
n = length(x2)
B = 1000
tau.boot <- rep(NA, B) # Init
for(b in 1:B){
x.boot <- rbeta(n, 1, psi(x2)) # bootstrap samples
tau.boot[b] <- 1/(1 + psi(x.boot))
}
bias = mean(tau.boot - tau(x2))
print("Boot bias")
bias
se = sd(tau.boot)
print("Boot se")
se
print("Asymptotical se")
se_tau(x2)
print("lower Ci")
lower = tau(x2) - qnorm(1 - alpha2/2)*se
lower
print("upper Ci")
upper = tau(x2) + qnorm(1 - alpha2/2)*se
upper
```
Non parametric Bootstrap
```{r}
tau.nboot <- rep(NA, B) # Init
for (b in 1:B){
idx <- sample(x = 1:n, replace = T)
x.boot <- x1[idx]
tau.nboot[b] <- 1/(1 + psi(x.boot))
}
bias = mean(tau.nboot - tau(x1))
bias
se = sd(tau.nboot)
se
lower = tau(x1) - qnorm(1 - alpha2/2)*se
lower
upper = tau(x1) + qnorm(1 - alpha2/2)*se
upper
```
# January 2018 Prob
```{r}
# Probs of selecting each district
di = c(1/3,1/3,1/3) # prob inside sample must be vector, not scalar
L = c(0.21,0.45,0.75) # same here

M <- 10000
dis <- sample(1:3, M, replace = T, prob = di)
# get the voters
lib_vec <- rbinom(M, 1, L[dis]) # 1 = Liberals, 0 = Others
mean(lib_vec)
# True is 0.47
```
```{r}
s =sqrt(20 - (16*0.68))
m = -8
1 - pnorm(0,m,s)
```
```{r}
p = 0.52
n = 990
z= qnorm(1 - 0.05/2)
se = sqrt(p * (1 - p) / n)
lower = p - z * se
lower
upper = p + z * se
upper
```

```{r}
x = c(-1, -1, 0)
teta_ml = sum(abs(x))/length(x)
teta_ml
tau_ml = sqrt(teta_ml)/2
tau_ml
# CI
se = sqrt(teta_ml * (1 - teta_ml)/length(x))
se
alpha = 0.1
z = qnorm(1-alpha/2)
lower = teta_ml - z*se
lower 
upper = teta_ml + z*se 
upper
```

```{r}
# Likehood func
x <- c(-1, -1, 0)
xx = c(-1, -1, 0)
L = function(theta, x){
  p1 <- (theta / 2)^sum(abs(x))  
  p2 <- (1 - theta)^(length(x) - sum( abs(x))) 
  return(p1 * p2)  
}

curve(L(x, xx), from = 0, to = 1)
segments(teta_ml, 0, teta_ml, L(x,xx))
```

```{r}
my.pmf= function(x, theta){
  if (theta <= 0 || theta >= 1) stop("theta must be between 0 and 1")  # Check theta
  if (x == -1) return((theta / 2))  # Case for x = -1
  if (x == 0) return(1 - theta)      # Case for x = 0
  if (x == 1) return((theta / 2))   # Case for x = 1
  return(0)  # If x is not -1, 0, or 1, return 0 since it's discrete
}

# Parametrio Bootstrap
n = length(x)
B = 1000
t.boot = rep(NA, B)
for (b in 1:B){
  x.boot = sample(c(-1,-1,0), n, replace = TRUE, prob = c(my.pmf(-1, teta_ml),my.pmf(0, teta_ml), my.pmf(1, teta_ml) ))
  t.boot[b] = sum(abs(x.boot))/length(x.boot)
}
bias = mean(t.boot) - teta_ml 
bias
se = sd(t.boot) # se
se
a = 0.1
z = qnorm(1-a/2)
lower = teta_ml - z*se
lower
upper = teta_ml + z*se
upper # the upper bound is larger that 1! we don't have enought data


```

```{r}
(1.9 - (0.64 + 0.72)) * 1/1.9
```
# Jan 2018 Stats
```{r}
data = c(10.2, 10.0, 8.0, 1.8, 6.5, 4.4, 9.3, 7.6, 10.1, 7.8)
psi = function(x){sqrt(sum(x^2)/(2*length(x)))}
psi_ml = psi(data)
psi_ml
tau = function(x) {(2 - pi/2)*psi_ml^2}
tau(data)
```
```{r}
to = 15
a = 0.05
z = qnorm(1 -a/2)
se_tau = function(tpsi_ml, x){
  abs(psi_ml) / (2 * sqrt(length(x)))
  
}
se_t = se_tau(psi_ml,data)
se_t
W = abs((tau(data) - to)/se_t)
W
W > z # we cannot reject null
```
```{r}
func = function(x, psi) { (x/psi^2)*exp(-x^2/(2*psi^2))}

rmy = function(n,psi) {psi * sqrt(-2*log(1-runif(n)))}
M = 10000 # n
# PSI = 1
hist(rmy(M, 1), prob = TRUE)
curve(func(x, 1), add = TRUE)

# PSI = 10
hist(rmy(M, 10), prob = TRUE)
curve(func(x, 10), add = TRUE)
```

```{r}
psi0 = sqrt(15 / (2 - pi/2)) #psi0
# Wald test
to = 15
n = 100
a = 0.05
M = 1000
matr = matrix(rmy(n*M, psi =psi0), nrow= M, ncol = n)
#matr
tau = function(psi){(2 - (pi/2))*psi^2}
psi = function(x){sqrt(sum(x^2)/(2*length(x)))}
se_tau = function(x){
  ( abs(2*(2 - pi/2)*mle(x)) )*(abs(mle(x)))/(2*sqrt(length(x)))
}
psi_m = apply(matr,1,psi)
se_t = apply(matr, 1, se_tau)
tau_m = tau(psi_m)
W = abs((tau_m - to)/se_t)
mean(abs(W) > z) # we cannot reject null
```
# Prob 2018 Exam 3 Feb 14
```{r}
marginals = c(A=2/9, B = 4/9, C = 2/9, D=1/9)
conditionals = c(A = 0.3, B = 0, C = 0.3, D = 0.6)
part = names(marginals)
approv = c("ok", "no ok")
M = 1000
sim_partner= character(M)
sim_approval= character(M)

sim = rep(NA, M)
for (i in 1:M){
  partner = sample(part, size = 1, prob = marginals)
  tullia = sample(approv, size = 1, prob = c(conditionals[partner], 1-conditionals[partner]))
  sim_partner[i] <- partner
  sim_approval[i] <- tullia
}
sim <- data.frame(partner = sim_partner, approval = sim_approval)
sum(sim$approval=="ok" & sim_partner=="A")/sum(sim$approval=="ok")
```
# Stats Feb 2018 

```{r}
mse_mm = function(n, theta){
  theta^2/(3*n)
}
mse_mle = function(n, theta) {
  2 * theta^2 / ((n+1)*(n+2))
}
# theta 0.1
curve(mse_mm(x, 0.1), col = "red", from = 1, to = 15)
curve(mse_mle(x, 0.1), col = "blue", add = TRUE)
# theta 1
curve(mse_mm(x, 1), col = "red", from = 1, to = 15)
curve(mse_mle(x, 1), col = "blue", add = TRUE)
# theta 10
curve(mse_mm(x, 10), col = "red", from = 1, to = 15)
curve(mse_mle(x, 10), col = "blue", add = TRUE)
# MLE always better
```

```{r}
se = sqrt((0.9*0.1/1000)+(0.85*0.15/1000)) 
se
print("Ci at 80%")
# Ci at 80%
a = 0.2
z = qnorm(1-a/2)
d = 0.9 - 0.85
lower = d - z*se
lower
upper = d + z*se
upper

print("Ci at 95%")
# Ci at 95%
a = 0.05
z = qnorm(1-a/2)
d = 0.9 - 0.85
lower = d - z*se
lower
upper = d + z*se
upper
```

```{r}
teta = 5
n = 10
r_tau = teta/2
tau = function(x) {max(x)/2}
M = 10000
sim = matrix(runif(n*M, 0, teta), nrow = M, ncol = n)
tau_ml = apply(sim,1,tau)
mse = mean((tau_ml - r_tau )^2)
mse
```

```{r}

sqrt(0.54/1.04)
```
# Stats 2 2019
```{r}
x <- c(102, 97, 142, 117, 122, 128, 101, 133, 120, 117)
mle = function(x) {sum(x)/length(x)}
mle(x)
```
# Prob 3 2019
```{r}
n = 10
m = 1/2
sigma = 1/12
var_th = n*(m - m^2) + n * (n-1)*sigma
M = 1000
Q = runif(M)
Sn = rbinom(M, n, Q)
var_th
var(Sn)
```
# Stats 3 2019

```{r}
x <- c(102, 97, 142, 117, 122, 128, 101, 133, 120, 117)
a = 0.1 # CI at 90%
z = qnorm( 1- a/2)
m = mean(x)
se = sqrt(var(x)/length(x))
l = m - z*se
l
u = m + z*se
u
```

```{r}
xx <- c(0.55, 0.45, 0.10, 0.15, 0.80, 0.65, 0.38, 0.63, 0.10, 0.03)
psi = function(x){
  -length(x)/(sum(log(1-x)))
}
psi_ml = psi(xx)

L = function(psi,x){
  p = prod(1-x)
  psi^(length(x)) * p^(psi-1)
}
curve(L(x,xx), from = 0, to = 5)
abline(v = psi_ml)
```

```{r}
tau = function(x){
  1/(1+psi(x))
}
tau_ml = tau(xx)
# par bootstrap
n = length(xx)
B = 1000
tau.boot = rep(NA, B)
for (i in 1:B){
  x_boot = rbeta(n, 1, psi_ml)
  tau.boot[i] = tau(x_boot)
}
mean(tau.boot) -  tau_ml
se = sd(tau.boot)
se
a = 0.1
z = qnorm( 1 - a/2)

l = tau(xx) - z*se
l
u = tau(xx) + z*se
u
```
```{r}
# non param bootstrap
tau.nboot = rep(NA, B)
for (i in 1:B){
  idx = sample(1:n, replace = TRUE)
  x_boot = xx[idx]
  tau.nboot[i] = tau(x_boot)
}

mean(tau.nboot) -  tau_ml
se = sd(tau.nboot)
se
a = 0.1
z = qnorm( 1 - a/2)

l = tau(xx) - z*se
l
u = tau(xx) + z*se
u
```
```{r}
(60 - 56.3)/20
```

