---
title: "SDS2"
author: "Emanuele Iaccarino"
date: "2025-05-07"
output: html_document
---

# Midterm 2021.1

### .B

In a Bayesian conjugate analysis for a Bernoulli trial, the larger the sample size, the smaller the influence of the prior mean elicitation on the posterior mean.

```{r}
#?rbinom
teta = 0.7
a = 2
b = 4
try = c(10,100,1000)
for (n in try)
{
  y = rbinom(n ,size = 1, prob = teta)
  y_sum = sum(y)
  y_mu = mean(y)
  p_a = a + y_sum
  p_b = b + n - y_sum
  p_mu = p_a / (p_a + p_b)
  pr_mean = a / (a+b)
  
  print(n)
  print(y_mu)
  print(p_mu)
  print(pr_mean)
}

# At small sample sizes , the posterior mean is closer to the prior.
# As sample size increases, the posterior mean moves closer to the sample mean, until they nearly match
```
So the answer is TRUE

### .D

```{r}
x_sim = rbinom(10000, size = 1, prob = 0.4) # Simulation from Bernoulli
# Simulation of \theta values
theta_sim = rnorm(n=10000,
    mean = x_sim * 2 + (1 - x_sim) * (-3), # for x = 1
    sd   = sqrt(5) * x_sim + sqrt(4) * (1 - x_sim)) # for x = 0
joint_vect_sim_matrix = cbind(x_sim, theta_sim) # matrix with two columns (x_i,\theta_i)
theta_given_x0  = joint_vect_sim_matrix[x_sim == 0, 2] # we take second column (ony \theta values) where x_i = 0
cat("Empirical mean:", mean(theta_given_x0), "\n")
cat("Empirical sd:", sd(theta_given_x0), "\n")

hist(theta_given_x0, probability = TRUE) # empirical distr
curve(dnorm(x, mean = -3, sd = 2), col = "red", add = TRUE) # theorical curve
```
This demostrate that It is the realization of i.i.d. random variables drawn from the conditional distribution of θ∣X=0
So the answer D is correct


### .E
```{r}
y_obs = c(3, 1, 1, 1, 1, 1, 1, 1, 1, 1, 0, 1, 2, 1, 3, 0, 2, 3, 0, 2, 1, 0, 3, 4, 1, 2, 0)
n = length(y_obs)
# Yi is distr as a Poisson, i.i.d. with \theta > 0
# Also we know from the problem
prior_mu = 2.1
prior_sd = 3

# We now have to obtait the prior gamma
# we solve a sistem of equation where
# \theta is distr as a gamma (a, b)
# and where
# E[\theta] = a/b = 2.1
# var[\theta] = a / b^2 = 9

# 2.1 * b / b^2 = 9 -> 2.1 / b = 9 - > 
b =  prior_mu / prior_sd^2
a = b * prior_mu

# P prior  \theta < 1
pgamma(1, shape = a, rate = b)
```


# Midterm 2021.2


### .A
```{r}
y_obs = c(0, 0, 0, 1, 1, 0, 0, 1, 1, 1, 1, 1)
success = sum(y_obs) 
n = length(y_obs)
failures = n - success
# Prior: Beta (4,5)
a = 4
b = 5
curve(dbeta(x,a,b))
# unimodal
```

```{r}
# Posterior distribution
a_post = a + success
b_post = b + failures
# Posterior median
print("Posterior median")
qbeta(0.5, a_post, b_post)
# 80% credible interval
print("80% credible interval")
alpha = 0.2
a_mezzi = alpha / 2
print("lower")
qbeta(a_mezzi, a_post, b_post)
print("upper")
qbeta(1 - a_mezzi, a_post, b_post)
```

```{r}
# Posterior probability that \theta > 0.4
print("Posterior probability that \theta > 0.4")
post_prob = 1 - pbeta(0.4, a_post, b_post)
post_prob
post_odds = post_prob / (1 - post_prob)

# Prior odds that \theta > 0.4
print("Prior odds that \theta > 0.4")
prior_prob = 1 - pbeta(0.4, a, b)
prior_odds = prior_prob / (1 - prior_prob)
# prior_prob
prior_odds
print("Bayesian Factor")
BF = post_odds / prior_odds
BF
```

### .B

```{r}
obs_succ = 12
p_succ = 0.45
# size = ? , parameter of the Binomial
N_max = 200
# Prior: uniform over all N between 12 and 200
N_vals = obs_succ:N_max # values from : to
prior = rep(1 / length(N_vals), length(N_vals))
# Rembember when asked to Retrieve values for N = 12

# Likelihood is the prob of having 12 successes given Nmax and p_succ = 
# (N on 12) * p_succ ^ obs_succ * (1 - p_succ)^ ( N_max - obs_succ)
# That more simply on R is:
print("likelihood n12")
likelihood = dbinom(obs_succ, size = N_vals, prob = p_succ)
likelihood_n12 = dbinom(obs_succ, size = 12, prob = p_succ)
likelihood_n12
#likelihood
# Posterior (unnormalized)
print("Posterior (unnormalized)")
unnormalized_post = likelihood * prior
#unnormalized_post
# Normalize
print(" Normalize posterior n12")
posterior = unnormalized_post / sum(unnormalized_post)
posterior_n12 = posterior[which(N_vals == 12)] # should be first index since we start counting from obs_succ on N_vals, btw let's keep this "general" approach
posterior_n12
# Posterior mean
print("Posterior mean")
post_mean = sum(N_vals * posterior)
post_mean
# Posterior mode (N with highest posterior probability)
print("Posterior mode (N with highest posterior probability)")
post_mode = N_vals[which.max(posterior)]
post_mode
# Posterior median
cumulative = cumsum(posterior)
print("Posterior median")
post_median = N_vals[which(cumulative >= 0.5)[1]]
post_median
```

### .C

```{r}
# Yi are distr as a N(mean = \theta = ? unknown , var = sigma^2 = 100)
# From table when Normal(θ, σ^2) and we have (known σ^2):
# θ ∼ Normal(µ0, τ0^2) as Posterior hyperparameter
sigma2  = 100
y_obs = c(
  68.443, 72.576, 84.826,
  69.255, 55.383, 68.454, 68.678,
  79.598, 75.484, 74.460, 77.024,
  79.574, 85.889, 63.586,
  76.228, 62.234, 82.880,
  75.368, 72.744, 71.131, 46.112,
  61.266, 70.219, 74.881,
  66.126, 63.556, 77.263, 67.144, 71.113, 64.156
)
# From now on there might be some error cause I didn't copied exactly the y_obs cause the pic was blurry and we talk about 10 decimals, let's say an approximation here is good
n = length(y_obs)
mu0 = 75
tau2   = 1.25
theta = seq(70, 80, length.out = 1000)

# Prior density
prior_density = dnorm(theta, mean = mu0, sd = sqrt(tau2))

# Plot
plot(theta, prior_density)
# tribution is simmetric
```

```{r}
y_mean = mean(y_obs)
# Posterior calculations
post_var = 1 / (n / sigma2 + 1 / tau2) # 1/k on the tables
post_sd = sqrt(post_var)
# It's the first formula
post_mean = post_var * ( mu0 / tau2 + sum(y_obs) / sigma2 ) 
print("Posterior mode")
post_mean # Since the distribution is simmetric post_mean = post_mode

# 95% credible interval for Normal
print("lower")
qnorm(0.025, post_mean,post_sd)
print("upper")
qnorm(0.975, post_mean,post_sd)
# Posterior probability \theta < x
x = 72.3296748
print("Posterior probability \theta < x")
posterior_prob = pnorm(x, mean = post_mean, sd = post_sd)
posterior_prob
# Prior probability \theta < x
print("Prior probability \theta < x")
prior_prob = pnorm(x, mean = mu0, sd = sqrt(tau2))
prior_prob
# Odds
posterior_odds = posterior_prob / (1 - posterior_prob)
prior_odds = prior_prob / (1 - prior_prob)
# Bayes Factor
print("Bayesian Factor")
BF = posterior_odds / prior_odds
BF
```
Is the conclusion based on the Bayes Factor in agreement with the one based on posterior probability?

Yes. even though the posterior probability of H0 is small, it's larger than expected under the prior, and that's why the Bayes Factor is > 1 in favor of H0.

# Midterm 2023

### .A

```{r}
prop_post_masses = c(0.000, 0.001, 0.003, 0.007, 0.012, 0.018, 
                      0.026, 0.034, 0.043, 0.051, 0.057, 0.062, 
                      0.065, 0.066, 0.065, 0.063)
theta_vals = 0:15
# Normalize Posterior
post_probs = prop_post_masses / sum(prop_post_masses)
print("Posterior mean")
post_mean = sum(theta_vals * post_probs)
post_mean
print("Posterior Median")
post_cdf = cumsum(post_probs)
post_median = theta_vals[which(post_cdf >= 0.5)[1]]
post_median
print("Posterior Mode")
post_mode = theta_vals[which.max(post_probs)]
post_mode
# 80% ET credible set
# we’re trimming 10% from each side
print("lower")
lower_bounds = theta_vals[which(post_cdf <= 0.1)]
#lower_bounds
lower_bound = tail(lower_bounds, 1) + 1 # +1 cause in R indexes start from 0
lower_bound
print("upper")
upper_bounds = theta_vals[which(post_cdf <= 0.9)]
# <= cause it has not to exceed 90%, that's why we can't use the method used for the lower bound
#upper_bounds
upper_bound = tail(upper_bounds, 1) + 1 # we take the largest value of this vector, that still belong to that central 80%
upper_bound # any larger value of θ would go beyond the 80% interval
```

I was told JUST NOW that we are allowed to use libraries, so here is the easy way to compute the intervals

```{r}
#install.packages("coda")
library(coda)
# Simulate posterior samples (large number = good approximation)
samples = sample(theta_vals, size = 100000, replace = TRUE, prob = post_probs)
samples_mcmc = as.mcmc(samples) # Convert to mcmc object
print("80% equal-tailed")
quantile(samples, probs = c(0.1, 0.9))  # for 80% equal-tailed
print("95% HPD")
HPDinterval(samples_mcmc, prob = 0.95) # for 95% HPD

```

```{r}
# unnormalized prior vector
prop_prior_masses = c(0.128, 0.112, 0.092, 0.073, 0.056, 0.042,
                       0.031, 0.022, 0.016, 0.011, 0.008, 0.005,
                       0.004, 0.002, 0.002, 0.001)

theta_vals = 0:15
# Normalize the prior
prior_probs = prop_prior_masses / sum(prop_prior_masses)
# Prior mean and sd
prior_mean = sum(theta_vals * prior_probs)
print("Prior Sd")
prior_sd = sqrt(sum((theta_vals - prior_mean)^2 * prior_probs))
prior_sd
# Compute the MLE of \theta
# Since it's based only on the prior (no data), MLE is not about likelihood
print("MLE of \theta")
theta_mle = theta_vals[which.max(prop_post_masses)] 
theta_mle # ? Review cause I think is wrong
```

### .B

```{r}
# conjugate normal-normal Bayesian model

y_obs = c(75.66, 89.76, 79.44, 93.03, 93.71, 98.32, 83.79, 77.93, 81.93, 56.91, 72.09, 88.15, 81.02, 85.68, 70.93)
n = length(y_obs)
y_bar = mean(y_obs)
sigma2 = 100
tau2 = 1.25
mu0 = 75
# plot tr
# Plot range
x_vals = seq(70, 80, length.out = 500)
prior_density = dnorm(x_vals, mean = mu0, sd = sqrt(tau2))

# Plot
plot(x_vals, prior_density)
# is symmetric
```

```{r}
# Posterior variance and mean (Normal-Normal conjugate formulas)
posterior_var = 1 / (1 / tau2 + n / sigma2 ) # 1/k on cheatsheet
posterior_sd = sqrt(posterior_var)
posterior_mean = posterior_var * ( mu0 / tau2 + sum(y_obs) / sigma2)

# Posterior mode = posterior mean in Normal case (distribution is symmetric)
posterior_mode = posterior_mean
print("Posterior Mode")
posterior_mode
```

```{r}
# 90% credible interval (equal-tailed since Normal is symmetric)
print("90% credible interval")
qnorm(0.05, posterior_mean, posterior_sd)
qnorm(0.95, posterior_mean, posterior_sd)
print("Posterior probability \theta < x")
x = 74.6258756
posterior_prob = pnorm(x, posterior_mean, posterior_sd)
posterior_prob
posterior_prob_greater = 1 - posterior_prob
posterior_odds = posterior_prob / posterior_prob_greater

print("Prior Odds \theta < x")
prior_prob_less = pnorm(x, mu0 , sqrt(tau2)) # P(\theta < x)
prior_prob_greater = 1 - prior_prob_less # P(\theta > x)
prior_odds = prior_prob_less / prior_prob_greater
prior_odds

print("Bayesian Factor")
BF = posterior_odds / prior_odds
BF
```

The posterior probability that \theta < x(74.6258756) is low (~0.077), which means we do not support H0.
The Bayes Factor BF is 0.1428431, much less than 1, indicating evidence in favor of H1 ( i.e., \theta > x(74.6258756) ).

### .C

```{r}
set.seed(5419)
y_obs <- rbinom(11, size = 1, prob = 0.75)
sum(y_obs)  # number of successes
# that’s a Beta(1,1) prior
# Posterior calculation
a_post <- 1 + sum(y_obs)
b_post <- 1 + (11 - sum(y_obs))
# 0.85 equal-tailed credible interval
# qbeta(c(0.075, 0.925), a_post, b_post)
# This is "wrong" cause it does not provide the SHORTEST CI

# Find the shortest credible interval for the posterior with 85% probability
# We are gonna try an empirical approach to reach the shortest CI
samples <- rbeta(1e6, shape1 = a_post, shape2 = b_post)

# Function 
get_HPD <- function(samples, cred_mass = 0.85) {
  sorted <- sort(samples)
  n <- length(sorted)
  interval_idx <- floor(cred_mass * n)
  widths <- sorted[(interval_idx+1):n] - sorted[1:(n - interval_idx)]
  min_idx <- which.min(widths)
  c(sorted[min_idx], sorted[min_idx + interval_idx])
}

get_HPD(samples, cred_mass = 0.85)

# Method used by the professor in his demos
# install.packages("TeachingDemos")
library(TeachingDemos)

# posterior distribution (Beta posterior in our case)
posterior_icdf <- function(p) {
  qbeta(p, shape1 = a_post, shape2 = b_post)
}

HPD_interval <- hpd(p = posterior_icdf, conf = 0.85)
HPD_interval
```

```{r}
# system of equations
target_mean <- 0.75 # \alpha / (\alpha + \Beta)
target_prob <- 0.95 # P(θ > 0.5)

# Use optimization
objective <- function(par) {
  # We compute how far α and β are from satisfying both constraints
  alpha <- par[1]
  beta <- par[2]
  
  mean_error <- (alpha / (alpha + beta) - target_mean)^2
  prob_error <- (1 - pbeta(0.5, alpha, beta) - target_prob)^2
  
  return(mean_error + prob_error) # we minimize the total squared error
}

# Initial guess: c(2, 1)
result <- optim(c(2, 1), objective, method = "L-BFGS-B", lower = c(0.1, 0.1))
# We search for values of α and β that make the total error as close to 0 as possible
# Extract solution
alpha <- result$par[1]
beta <- result$par[2]
alpha
beta
```
### .D

```{r}
# Observed failure times
y <- c(0.6, 4.6, 4.6, 2.2, 39.0, 0.5, 14.5, 7.0, 4.1, 
       5.5, 2.8, 12.2, 22.4, 6.9, 8.5, 57.8, 12.1, 13.9, 
       21.3, 19.4)
n = length(y)
# Prior parameters
alpha_prior <- 20 # rate
beta_prior <- 20 # shape
# Posterior parameters
# Gamma(a + y, b + n)
alpha_post <- alpha_prior + n
beta_post <- beta_prior + sum(y)
# Median of posterior distribution (Gamma)
# REMEMBER: shape goes first, than rate
lambda_median <- qgamma(0.5, beta_post, alpha_post)
# median of the posterior \theta is reciprocal of the expected failure time (since the data is modeled as exponential with rate \theta)ù

# median of θ = 1/λ
theta_median <- 1 / lambda_median
print("median of θ = 1/λ")
theta_median
```

# Midterm 2024.1

### .A

















# [2024] Quizzes on conjugate Bayesian analysis of discrete data models (Bernoulli, binomial, Poisson)

Suppose we have observed the following 14 conditionally i.i.d. Bernoulli random variables with unknown success probability \theta (theta)

y_obs

##  [1] 0 1 0 0 0 0 0 0 0 0 0 0 0 0

and we want to carry out some conjugate Bayesian analysis using as a tribution a Beta(4,4)

The tribution is

```{r}
y_obs = c(0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
n = length(y_obs)
success = 1
failure = n-success
a = 4
b = 4
# plot prio distr
curve(dbeta(x,a,b))
# answer is unimodal
# no flat, it would have been fir beta (1,1)
# not wrong surely
# not improper cause it integrate to 1
```

```{r}
# Yi | \theta distr as a Bernoulli (\theta)
# Prior: \theta distr Beta (a=4,b=4)
pr_mean = a / (a+b)
pr_var = (a*b) / ( (a+b)^2 * (a+b+1) )
print("Prior mean and var")
pr_mean
pr_var

# Posterior: \theta distr Beta(a + y, b + n - y) where y = success
# \theta distr Beta(4+1, 4+14-1) = Beta (5,17)
a_p = 5 # a + succ
b_p = 17 # b + failures
post_mean = a_p / (a_p + b_p)
post_var = (a_p * b_p) / ( (a_p + b_p)^2 * (a_p + b_p +1) ) 
print("Posterior mean and var")
post_mean
post_var

# Posterior median
print("Posterior median")
qbeta(0.5, a_p, b_p)
# Confidence interval at 90%
alpha = 0.1
a_half = alpha/2
print("lower")
qbeta(a_half, a_p, b_p)
print("upper")
qbeta(1 - a_half, a_p, b_p)
# Posterior probability that \theta > 0.4
p = 0.4
print("Posterior probability that \theta > 0.4")
post_prob = 1 - pbeta(p, a_p, b_p)
post_prob
post_odds = post_prob / (1 - post_prob)
# Prior odds for \theta > 0.4
print("Prior odds for \theta > 0.4")
prior_prob = 1 - pbeta(p, a, b)
a
prior_odds = prior_prob / (1-prior_prob)
prior_odds
#Bayesian Factor
BF = post_prob / prior_odds
print("Bayesian Factor")
BF
```

Suppose we have observed the following counts of daily complaints received at 30 different stores of the same food supermakets.

```{r}
y_obs = c(12, 3, 7, 11, 7, 7, 12, 7, 8, 8, 4, 11, 5, 3, 6, 12, 11, 9, 8, 15, 8, 9, 7, 8, 6, 4, 12, 11, 9, 9)
```

Suppose we have some prior opinion on the expected value of the number of complaints which is represented by a Gamma distribution with rate parameter 0.8 and shape parameter 2.
```{r}
n = length(y_obs)
sum_y = sum(y_obs)
# Prior: \theta distr as a Gamma (a = shape = 2, b = rate = 0.8)
a = 2
b = 0.8
```

Consider as a suitable statistical model for non negative counts the Poisson distribution and compute the median of the posterior distribution.

```{r}
# We model as a Poisson distr
# Posterior: \theta | y_obs distr as a Gamma (a + n, b + sum_y)
a_p = a + sum_y
b_p = b + n

post_mean = a_p / b_p
post_median = qgamma(0.5, a_p, b_p)
post_mode = (a_p - 1 ) / b_p
post_median
```
# [2024] Quizzes on simulation, MC approximations and predictive distributions [**NEW**]

Use the following R code to provide a Monte Carlo approximation of the expectation of the posterior predictive distribution for the future observable Y_{21}

```{r}
# R code for approximating Bayesian inference via Monte Carlo approach
# conjugate Poisson model
set.seed(123)
theta_star = 4
y_obs = rpois(20,lambda = theta_star)
y_obs
##  [1] 3 6 3 6 7 1 4 7 4 4 8 4 5 4 2 7 3 1 3 8
mcsize = 100000
# prior = teta
prior_sim = rgamma(mcsize,shape=2,rate=2) # gamma (shape, rate)
# posterior = teta|X
posterior_sim=rgamma(mcsize,shape=2+sum(y_obs),rate=2+20) # gamma (shape + n(X), rate + n)
```

```{r}
hist(prior_sim, col = "red") # asymm right tail distr
hist(posterior_sim, add = TRUE, col = "blue") # normal distr
```

```{r}
y21_sim = rpois(mcsize, lambda = posterior_sim)
# Approximate E[Y_21 | y_obs]
mean(y21_sim)
```